Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-08 paper/intro.tex
easier reviews

\begin{figure}
    \centering 
    \subfloat{
        \includegraphics[width=0.475\linewidth]{intuition.png}
    }
    \subfloat{
        \begin{tikzpicture}
            \node (push) at (-2,0) {};
            \node [circle, draw=red, thick] (a) at (0,4) {A};
            \node [circle, draw=green, thick] (b) at (4,4) {B};
            \node [circle, draw=blue,  thick] (c) at (0,2) {C};
            \node [circle, draw=red, thick] (d) at (2,2) {D};
            \node [circle, draw=green, thick] (e) at (4,2) {E};
            \node [circle, draw=blue, thick] (f) at (0,0) {F};
            \node [circle, draw=red, thick] (g) at (2,0) {G};
            \node [circle, draw=green, thick] (h) at (4,0) {H};
            \draw (a) -- (c);
            \draw (a) -- (d);
            \draw (b) -- (e);
            \draw (c) -- (f);
            \draw (d) -- (f);
            \draw (d) -- (g);
            \draw (e) -- (h);
        \end{tikzpicture}
    } \caption{Network based spatio-temporal data representation. \textbf{Left}:
    Three temporal stages of the evolution of a fictitious urban area, with
    aggregation areas A to H. \textbf{Right}: Network representation of the
    aggregation areas where the colours identify similar regions.
        \label{fig:intuition}}
\end{figure}

Instead, we propose a network representation. Intuitively, a \emph{network}
(also called a \emph{graph}) is a collection of entities (nodes) that are
related to each other (edges). In this case, each different aggregation area is
represented as a node and we connect nodes that have overlapping geographical
areas, leading to the network illustrated on the right of
Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
that are similar, we are effectively identifying clusters in the spatio-temporal
data, as illustrated by the colours of the nodes on the right side of
Figure~\ref{fig:intuition}. Further, the possible evolution paths can also be
obtained by computing temporal sequences of nodes, in this case: (A, C, F), (A,
D, F), (A, D, G), and (B, E, H). This representation is also suited for
geographically consistent regions, as illustrated by the stable park in this
example, and is therefore a generalisation of the traditional paradigm.

>>>>

\begin{figure}
    \centering 
    \subfloat{
        \includegraphics[width=0.475\linewidth]{intuition.png}
    }
    \subfloat{
        \begin{tikzpicture}
            \node (push) at (-2,0) {};
            \node [circle, draw=red, thick] (a) at (0,4) {A};
            \node [circle, draw=green, thick] (b) at (4,4) {B};
            \node [circle, draw=blue,  thick] (c) at (0,2) {C};
            \node [circle, draw=red, thick] (d) at (2,2) {D};
            \node [circle, draw=green, thick] (e) at (4,2) {E};
            \node [circle, draw=blue, thick] (f) at (0,0) {F};
            \node [circle, draw=red, thick] (g) at (2,0) {G};
            \node [circle, draw=green, thick] (h) at (4,0) {H};
            \draw (a) -- (c);
            \draw (a) -- (b);
            \draw (a) -- (d);
            \draw (b) -- (e);
            \draw (c) -- (f);
            \draw (c) -- (d);
            \draw (d) -- (f);
            \draw (d) -- (e);
            \draw (d) -- (g);
            \draw (e) -- (h);
            \draw (f) -- (g);
            \draw (g) -- (h);
        \end{tikzpicture}
    } \caption{Network based spatio-temporal data representation. \textbf{Left}:
    Three temporal stages of the evolution of a fictitious urban area, with
    aggregation areas A to H. \textbf{Right}: Network representation of the
    aggregation areas where the colours identify similar regions.
        \label{fig:intuition}}
\end{figure}

Instead, we propose a network representation. Intuitively, a \emph{network}
(also called a \emph{graph}) is a collection of entities (nodes) that are
related to each other (edges). In this case, each different aggregation area is
represented as a node and we connect nodes that have overlapping geographical
areas, leading to the network illustrated on the right of
Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
that are similar, we are effectively identifying clusters in the spatio-temporal
data, as illustrated by the colours of the nodes on the right side of
Figure~\ref{fig:intuition}. Further, all the possible paths of change can be
obtained by computing sequences of nodes over time, in this case: (A, C, F), (A,
D, F), (A, D, G), and (B, E, H). This representation is also suited for
geographically consistent regions, as illustrated by the stable park in this
example, and is therefore a generalisation of the traditional paradigm.

### 2019-08-08 paper/method.tex
GEAN SI submitted version
easier reviews

\subsection{Census methodology and data representation}
\info{I removed any specifics of the experiments (CTs, which variables) from here because this section is theory}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and there is geographical overlap between them. This
approach leads to a single network representing the whole spatio-temporal space
of the data. Our objective then becomes to identify partitions of this network
such that the nodes of each partition are more similar between themselves than
to the other nodes.

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. We introduced one new
parameter to this method: a maximum distance threshold for the merges, to avoid
the early merging of outliers. We refer the reader to the original
paper~\citep{markus2017} for more details, including a complete performance
evaluation using several metrics.

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
interquantile ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concept of \emph{trajectories}. Trajectories are composed by
regions classified into the same sequence of clusters over the considered
period. This enables direct access to regions that evolved in the same manner.
While individual census tracts remain interesting, these trajectories are the
main unit of exploration in this work.

>>>>

\subsection{Census methodology and data representation}

@@

Each area of each census year is represented as a node, and edges are placed
between nodes if the corresponding regions share geographic borders in the same
year. Further, edges are placed between nodes if the corresponding regions
belong to sequential years and \revision{their geometries intersect. To avoid
spurious connections caused by geometry fluctuations, one of the geometries is
slightly shrunk before the intersection, using a buffer of -1e6. More
importantly, while weights will be associated with these edges before they are
processed, they are not derived from the geometry, but from the data. The actual
intersection area is not considered in this representation. } This approach
leads to a single network representing the whole spatio-temporal space of the
data. Our objective then becomes to identify partitions of this network such
that the nodes of each partition are more similar between themselves than to the
other nodes.

@@

Once the distances are associated to the edges, we use watershed
cuts~\citep{Cousty2009} to create an initial clustering, which is then refined
into a hierarchy using the Sorted Maximal Matching (SMM)~\citep{markus2017} with
median linkage. The initial watershed step is performed to create an initial
clustering and reduce the running time of the SMM. We introduced one new
parameter to this method: a maximum distance threshold for the merges, to avoid
the early merging of outliers. We refer the reader to the original
paper~\citep{markus2017} for more details, including a complete performance
evaluation using several metrics. \revision{We chose this algorithm because it
is simple and easily customisable, and while different algorithms will lead to
different results, our methodology should work with any hierarchical clustering
algorithm.}

Each resulting cluster is contiguous in the network. This means that two
similar, but non-contiguous, sets of areas will be classified into two different
clusters, which can be counter-intuitive. To overcome this issue, we
\emph{augment} the network with two new edges per node from a nearest neighbours
graph~\citep{scikit-learn} using only the distances between the histograms.
These edges connect nodes with similar content, if they are not already
connected, providing a path for the algorithm to group similar nodes.
\revision{The regions connected by those edges will be merged on the first
stages of the clustering, since they are very similar, leaving the remaining
steps of the hierarchy to be determined only by the geographical edges. We
explored with different numbers of augmentation edges, but the results were not
consistent, since the distribution of the edges is data dependent. Adding two
edges per node was the least number of edges that led to stable and consistent
results in the scenarios available in our prototype. Since the problem of
balancing the data space with the geographical space is relevant for
geographical data analysis, this methodology potentially warrants further
exploration, beyond the scope of this work.}

@@

We also use these statistical measurements to discover what characterises each
cluster, that is, what makes it different from the others.  We define the
\emph{relevance} of a variable of an aspect based on the distance between the
interquartile ranges (IQR) of the clusters in the same hierarchical level. If
the IQRs overlap for all clusters, that variable is not relevant to the
characterisation of the cluster, but if the IQRs are distant, it means that this
specific range of values is something that only occurs in this cluster.
Examining IQRs therefore provides users a straightforward visual method for
determining what variables most clearly define a given cluster.

\subsection{Clusters and trajectories}
While the partition of the data into different clusters helps the user to
understand what groups exist and where they are, we are also interested in the
evolution of these groups. To examine this process of evolution directly, we
introduce the concepts\revision{of \emph{temporal paths} and \emph{trajectories}.

We call a temporal path any sequence of nodes in our representation network such
that the temporal information associated with the nodes only increases. For
instance, in Figure~\ref{fig:intuition}, the sequences ACF, ADF, ADG, and BEH
are temporal paths. With harmonised data, the time-series of to each region
would form a temporal path, each node would be connected only to its older and
newer versions, belonging to only one temporal path, as illustrated by the path
BEF. Since our data is not harmonised, more connections are allowed and each
node can belong to an arbitrary number of paths.

Semantically, this is a generalisation of the idea of geographical time-series,
because each temporal path is one possible option for the data to change over
time. Returning to Figure~\ref{fig:intuition}, the paths ACF, ADF, and ADG all
start on the same homogeneous region, but evolve differently over time. In other
words, this network encodes the information that portions of the region A
changed to form regions C and D, but we do not know specifically which parts,
nor we need to, since the same region can belong to several temporal paths.
Interestingly, when interpreted in this framework, geographical harmonisation is
a method to split and/or merge nodes so each belongs to a single temporal path.

Since each node in the sequence that forms the temporal paths has an associated
cluster, we can classify the paths based on the sequence of clusters. We call
each unique sequence of clusters present in this result a \emph{trajectory}.
Regions on the same trajectory had the same sequence of clusters, therefore had
similar temporal evolution.}

### 2019-08-08 paper/related.tex
easier reviews

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014}. Despite its inherent
errors~\citep{Logan2016,Hallisey2017}, this dataset has become the standard
source for longitudinal demographic data at the neighbourhood scale, with
similar efforts appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}.
These datasets have been highly significant for the field. Yet they also limit
the universe of  data that can be used to study neighbourhood change, since any
new datasets would need to be similarly processed in order to be rendered
compatible with these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
increased spatial accuracy, this approach does not require complex harmonisation
when new data is considered, if the grids are compatible. However, demographic
data is usually not available in this format, especially from older sources.
Additionally, the conversion from tabulation areas can introduce significant
errors.

>>>>

\subsection{Longitudinal demographic studies}
Census data is used not only to discover demographic
patterns~\citep{Firebaugh2016}, but to correlate demographic characteristics to
other measurements~\citep{diez1997neighborhood}. However, longitudinal studies
are rare, because they are difficult : \emph{"(...) One of the most challenging
and fascinating areas in spatial statistics is the synthesis of spatial data
collected at different spatial scales(...)"}~\citep{gotway2002combining}. While
census tract level data is readily available for the US since at least
1910~\citep{nhgis}, most studies consider the period between 1970 and 2010,
using pre-harmonised data from the Longitudinal Tract Data
Base~\citep{Logan2014}. Despite its inherent
errors~\citep{Logan2016,Hallisey2017}, this dataset has become \revision{widely
adopted, along with the Neighborhood Change Database~\cite{ncdb},} as source for
longitudinal demographic data at the neighbourhood scale, with similar efforts
appearing in other countries~\citep{Liu2015,Lee2015,Allen2018}. These datasets
have been highly significant for the field. Yet they also limit the universe of
data that can be used to study neighbourhood change, since any new datasets
would need to be similarly processed in order to be rendered compatible with
these sources.

Another option considers the use of grid
data~\citep{Dmowska2017,Dmowska2018,stepinski2019imperfect}. Beyond the
potentially increased spatial precision, this approach does not require complex
harmonisation when new data is considered, if the grids are compatible. However,
demographic data is usually not available in this format, especially from older
sources. Additionally, the conversion from tabulation areas can introduce
significant errors.

### 2019-08-08 paper/study.tex
easier reviews

In this section we present three illustrative scenarios, using decennial census
data from the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010. The prototype interface allows access to 41 regions, 29
in the US and 12 in Canada. New York City was split into its boroughs to avoid
memory crashes on the client browser due to the high number of CTs.  We used
five aspects for the USA: Education level, Family income, Marital status,
Occupation, and Race; and seven for Canada: Age, Education level, Home language,
Household Income, Marital status, Occupation, Place of birth, and Religion.

@@

We consider a region that is approximately the administrative border of the
current city of Toronto, using all seven available aspects with equal weights.
While Chicago was fairly stable, Toronto is known to be a more dynamic and
diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially Asian~\citep{Fong2003}. 
Toronto is also known for a stable and well defined Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect the combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

\subsection{Los Angeles}

We selected for this scenario a region around the metropolitan area of Los
Angeles (LA), following urban density. While Chicago and Toronto demonstrated
the abilities of our method to examine stable and dynamic cities, LA is considerably
larger, both in population and area. Since our method identifies only the eight
most distinct groups, we expect that some groups will not be identified,
especially if their characterisation is similar to another groups. Further, our
system does not include the census variables related to Hispanic heritage, which
were only included in the newer censuses. However, from the literature, we
expect to see some clusters where Race is an important aspect, including White,
Black, and Asian~\citep{Reibel2003}.

The summary of the results using all aspects with equal weights and eight
clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
relevant aspect of each cluster is highlighted. From the trajectories plot, we
can see that there is a large but shrinking cluster, depicted in green, one
increasing cluster in purple, an almost constant orange cluster, a smaller but
increasing pink cluster, and three other small clusters. The corresponding map
illustrates where these clusters are located, and that they are somewhat
geographically stable, with some movement between the green, orange, and purple
clusters.

\begin{figure}
    \centering 
    \includegraphics[width=0.975\linewidth]{LA.pdf}
    \caption{Result for Los Angeles with 8 clusters, including three small and
        ephemeral clusters. Cluster characterisation is displayed
        in Figure~\ref{fig:laAll}.\label{fig:la}}
\end{figure}

From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
by a high percentage of White population, low percentage of population in the
lowest income bracket, mostly administrative occupations, and about 30\% of the
population with four or more years of college. The orange cluster is
characterised by a high percentage of Black population, with few people in the
highest range of income and education. The purple cluster corresponds to a high
concentration of "Other" in race, which includes Hispanic for this dataset, high
concentration of Labourers, and low education and income. The pink cluster
contains a high percentage of Asian population and about 30\% of the population
with four or more years of college. The light green cluster contains very few
people in the lower income bracket, mostly White population, with the highest
percentage of population with four or more years of college, working
administrative jobs, and a high concentration of singles. The yellow cluster
represents Black population, with higher level of education and income, mostly
working administrative jobs. The brown cluster represent a majority of single
population, working administrative jobs with mostly low income. The dark grey
cluster is characterised by all its population in the lowest income bracket, low
education level, with a majority of White population. Since the extremes in the
boxplots of the grey cluster are not significantly different, we can also
surmise that this cluster is either small or homogeneous.

\begin{figure*}
    \centering 
    \includegraphics[width=0.95\linewidth]{laAll.pdf}
    \caption{Full characterisation of the eight clusters found for LA. The red
     rectangles indicate the most relevant aspects for each
     cluster.\label{fig:laAll}}
\end{figure*}

The  green, orange, and purple clusters present a significant intra-cluster
variance in most variables, as indicated by extreme whiskers of the boxplots.
While fifty percent of the CTs in the green cluster have between 20\% and 40\%
of people in the lowest income bracket, that cluster also includes CTs where
none and all the population belongs to that bracket. This might indicate that
this cluster represents different groups of people that are not different enough
to be separated at this level of the hierarchy. Conversely, the light green,
brown, and dark grey clusters are different enough to be separated into their
own clusters at this level, despite being small and ephemeral, including only a
few CTs.

The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
the presence of green and purple tones mixed with the orange, which may warrant
further exploration. By clicking on the orange bar above the boxplot, we select
all trajectories that contain the orange cluster. The corresponding details are
illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
orange cluster is progressively replaced by the purple cluster on its east side,
and in turn expanding to the west. Interestingly, the population increased, the
racial profile changed, but the distribution of income was reasonably stable,
with a higher amount of the population in the lowest income range and very few
people in the highest income range. Indeed, the income  difference is
significant when compared to the city-wide distribution.

\begin{figure}
    \centering 
    \includegraphics[width=0.75\linewidth]{laBH.pdf}
    \caption{Top: Geographic changes in the majority Black population cluster
    (orange) and Labourers cluster (green). Bottom: Income evolution for this
    region (black) and the whole city (grey).\label{fig:labh}}
\end{figure}

A portion of this region is classified into the green cluster in 2010,
indicating a majority white population. To further understand that change, we
clear the current selection, and select all regions that changed from orange in
1970 to green in 2010, using the transition matrix. A portion of the resulting
region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
along with the temporal evolution of Race. Despite this difference, the other
aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
slightly lower income and education profiles. While the racial aspect changed
considerably, the economic and educational aspects stayed the same.

\begin{figure}
    \centering 
    \includegraphics[width=0.75\linewidth]{laVol.pdf}
    \caption{Details for a volatile region contained in the area of
         Figure~\ref{fig:labh}. This region went from Black to Hispanic to
        White.\label{fig:laVol}}
\end{figure}

These results are visually similar to the analysis presented by
Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
portions of the 'stable elite' cluster, near Inglewood, correspond to the high
income, high education yellow cluster. The other groups identified by that study
were clumped by our system into the green cluster, which would be further
divided if more than eight clusters were to be used.

If Toronto illustrates a relatively smooth process of whole-sale
diversification, Los Angeles shows the ability of our system and method to
uncover more intensive smaller-scale processes of neighbourhood change. This
capacity to examine and classify different types of change operating at
different scales demonstrates the flexibility of the method and tool for
advancing the more open-ended notion of "neighbourhood" envisioned by recent
neighbourhood research.

@@

Most of the experts demonstrated interest in using our method in their research
(A, B, D, E), aiming to use the census data as a backdrop for other datasets,
providing demographic context. They also mentioned the need to export subsets of
data, plots, and maps to be used in reports and publications (C, D, E). More
importantly, while these experts were aware that our method does not perform
geographic harmonisation, none of them mention it. We did not specifically ask
if this difference led to unexpected results, but rather if they found
interesting insights. Most experts found phenomena corroborated by the
specialised literature, indicating that our methodology produces equivalent
results, with a fraction of the effort. We interpret the fact that most of them
were interested in the next steps as confirmation of the accuracy of the method.

>>>>

In this section we present two illustrative scenarios, using decennial census
data from the United States~\citep{nhgis} and
Canada\footnote{\url{http://datacentre.chass.utoronto.ca/census/}}, tabulated by
CTs,  from 1970 to 2010. \revision{Accidentaly, the US data for 2010 is actually
not from the decennial census, but from the ACS 2006-2010. Further, the regions
selected do not correspond to any pre-defined regions (metros, cities, census
areas), but to arbitrary regions defined around a location of interest. We
selected a minimal set of aspects for each country, and they are not similar to
each other, out of convenience, since the variable matching was a manual
process. These caveats would likely compromise any serious attempt on a
comprehensive demographic study, but this is not the objective of this
experiment. We aim only to test and validate our methodology, comparing our
results to pre-existing literature insights. Indeed, our results were perfectly
aligned to several other studies, despite these methodological missteps, which
could arguably be an indicator of the robustness of our method.}

The prototype interface allows access to 41 regions, 29
in the US and 12 in Canada. New York City was split into its boroughs to avoid
memory crashes on the client browser due to the high number of CTs.  We used
five aspects for the USA: Education level, Family income, Marital status,
Occupation, and Race; and seven for Canada: Age, Education level, Home language,
Household Income, Marital status, Occupation, Place of birth, and Religion.

@@

We consider a region that is approximately the administrative border of the
current city of Toronto, using all seven available aspects with equal weights.
While Chicago was fairly stable, Toronto is known to be a more dynamic and
diverse city, with significant and increasing immigrant
population~\citep{hulchanski2007three,Fong2011}, especially
Asian~\citep{Fong2003}. Toronto is also known for a stable and well defined
Jewish community~\citep{Harold2018, Fong2011}. Therefore, we expect a
combination of stable and dynamic regions on the results, with Place of Birth,
Home Language, and Religion identified as relevant aspects. The results are
summarised in Figure~\ref{fig:to}, considering eight clusters.

@@

% \subsection{Los Angeles}

% We selected for this scenario a region around the metropolitan area of Los
% Angeles (LA), following urban density. While Chicago and Toronto demonstrated
% the abilities of our method to examine stable and dynamic cities, LA is considerably
% larger, both in population and area. Since our method identifies only the eight
% most distinct groups, we expect that some groups will not be identified,
% especially if their characterisation is similar to another groups. Further, our
% system does not include the census variables related to Hispanic heritage, which
% were only included in the newer censuses. However, from the literature, we
% expect to see some clusters where Race is an important aspect, including White,
% Black, and Asian~\citep{Reibel2003}.

% The summary of the results using all aspects with equal weights and eight
% clusters is illustrated in Figure~\ref{fig:la}. The full statistical description
% of the clusters is illustrated in Figure~\ref{fig:laAll}, where the most
% relevant aspect of each cluster is highlighted. From the trajectories plot, we
% can see that there is a large but shrinking cluster, depicted in green, one
% increasing cluster in purple, an almost constant orange cluster, a smaller but
% increasing pink cluster, and three other small clusters. The corresponding map
% illustrates where these clusters are located, and that they are somewhat
% geographically stable, with some movement between the green, orange, and purple
% clusters.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.975\linewidth]{LA.pdf}
%     \caption{Result for Los Angeles with 8 clusters, including three small and
%         ephemeral clusters. Cluster characterisation is displayed
%         in Figure~\ref{fig:laAll}.\label{fig:la}}
% \end{figure}

% From Figure~\ref{fig:laAll}, we can see that the green cluster is characterised
% by a high percentage of White population, low percentage of population in the
% lowest income bracket, mostly administrative occupations, and about 30\% of the
% population with four or more years of college. The orange cluster is
% characterised by a high percentage of Black population, with few people in the
% highest range of income and education. The purple cluster corresponds to a high
% concentration of "Other" in race, which includes Hispanic for this dataset, high
% concentration of Labourers, and low education and income. The pink cluster
% contains a high percentage of Asian population and about 30\% of the population
% with four or more years of college. The light green cluster contains very few
% people in the lower income bracket, mostly White population, with the highest
% percentage of population with four or more years of college, working
% administrative jobs, and a high concentration of singles. The yellow cluster
% represents Black population, with higher level of education and income, mostly
% working administrative jobs. The brown cluster represent a majority of single
% population, working administrative jobs with mostly low income. The dark grey
% cluster is characterised by all its population in the lowest income bracket, low
% education level, with a majority of White population. Since the extremes in the
% boxplots of the grey cluster are not significantly different, we can also
% surmise that this cluster is either small or homogeneous.

% \begin{figure*}
%     \centering 
%     \includegraphics[width=0.95\linewidth]{laAll.pdf}
%     \caption{Full characterisation of the eight clusters found for LA. The red
%      rectangles indicate the most relevant aspects for each
%      cluster.\label{fig:laAll}}
% \end{figure*}

% The  green, orange, and purple clusters present a significant intra-cluster
% variance in most variables, as indicated by extreme whiskers of the boxplots.
% While fifty percent of the CTs in the green cluster have between 20\% and 40\%
% of people in the lowest income bracket, that cluster also includes CTs where
% none and all the population belongs to that bracket. This might indicate that
% this cluster represents different groups of people that are not different enough
% to be separated at this level of the hierarchy. Conversely, the light green,
% brown, and dark grey clusters are different enough to be separated into their
% own clusters at this level, despite being small and ephemeral, including only a
% few CTs.

% The orange area in the map in Figure~\ref{fig:la} presents movement, indicated by
% the presence of green and purple tones mixed with the orange, which may warrant
% further exploration. By clicking on the orange bar above the boxplot, we select
% all trajectories that contain the orange cluster. The corresponding details are
% illustrated in Figure~\ref{fig:labh}. This shows a location change, where the
% orange cluster is progressively replaced by the purple cluster on its east side,
% and in turn expanding to the west. Interestingly, the population increased, the
% racial profile changed, but the distribution of income was reasonably stable,
% with a higher amount of the population in the lowest income range and very few
% people in the highest income range. Indeed, the income  difference is
% significant when compared to the city-wide distribution.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.75\linewidth]{laBH.pdf}
%     \caption{Top: Geographic changes in the majority Black population cluster
%     (orange) and Labourers cluster (green). Bottom: Income evolution for this
%     region (black) and the whole city (grey).\label{fig:labh}}
% \end{figure}

% A portion of this region is classified into the green cluster in 2010,
% indicating a majority white population. To further understand that change, we
% clear the current selection, and select all regions that changed from orange in
% 1970 to green in 2010, using the transition matrix. A portion of the resulting
% region, near the Florence-Graham region, is depicted in Figure~\ref{fig:laVol},
% along with the temporal evolution of Race. Despite this difference, the other
% aspects are similar to the ones from the region in Figure~\ref{fig:labh}, with
% slightly lower income and education profiles. While the racial aspect changed
% considerably, the economic and educational aspects stayed the same.

% \begin{figure}
%     \centering 
%     \includegraphics[width=0.75\linewidth]{laVol.pdf}
%     \caption{Details for a volatile region contained in the area of
%          Figure~\ref{fig:labh}. This region went from Black to Hispanic to
%         White.\label{fig:laVol}}
% \end{figure}

% These results are visually similar to the analysis presented by
% Dellmelle~\citep{Delmelle2016}.  What was there identified as 'persistently
% struggling' corresponds to the purple cluster in Figure~\ref{fig:la}; some
% portions of the 'stable elite' cluster, near Inglewood, correspond to the high
% income, high education yellow cluster. The other groups identified by that study
% were clumped by our system into the green cluster, which would be further
% divided if more than eight clusters were to be used.

% If Toronto illustrates a relatively smooth process of whole-sale
% diversification, Los Angeles shows the ability of our system and method to
% uncover more intensive smaller-scale processes of neighbourhood change. This
% capacity to examine and classify different types of change operating at
% different scales demonstrates the flexibility of the method and tool for
% advancing the more open-ended notion of "neighbourhood" envisioned by recent
% neighbourhood research.

@@

Most of the experts demonstrated interest in using our method in their research
(A, B, D, E), aiming to use the census data as a backdrop for other datasets,
providing demographic context. They also mentioned the need to export subsets of
data, plots, and maps to be used in reports and publications (C, D, E). More
importantly, while these experts were aware that our method does not perform
geographic harmonisation, none of them mention it. We did not specifically ask
if this difference led to unexpected results, but rather if they found
interesting insights.
